Numerical simulation of the Cordilleran ice sheet \\ through the last glacial cycle

### 2014-06-24 kappa.tex
Improve header visibility in LaTeX source.
Add macros to formalise notations.

% kappa.tex
% ----------------------------------------------------------------------

@@

% My commands
%\newcommand{\note}[1]{\textbf{[NOTE: #1]}}
\newcommand{\todo}[1]{\emph{[\textbf{Todo:} #1]}}
%\newcommand{\aref}[0]{\textbf{[ref.]}}
\renewcommand{\div}[1]{\mathrm{div}\,#1}
\renewcommand{\grad}[1]{\vec{\mathrm{grad}}\,#1}
\newcommand{\tensdiv}[1]{\vec{\mathrm{div}}\,#1}
\newcommand{\tensgrad}[1]{\bm{\mathrm{grad}}\,#1}

@@

% ----------------------------------------------------------------------
\begin{document}
% ----------------------------------------------------------------------

@@

% ----------------------------------------------------------------------
\section{Introduction}
\label{sec:intro}
% ----------------------------------------------------------------------
% Ice as viscous matter
% Glaciers and ice sheets
% The last glacial cycle

@@

% ----------------------------------------------------------------------
\section{Field area description}
% ----------------------------------------------------------------------
% Geographic setting
% Climatic setting
% Palaeo-glaciology

@@

% ----------------------------------------------------------------------
\section{Numerical ice sheet model}
% ----------------------------------------------------------------------

\subsection{Overview}

@@

\subsection{Field equations}

The thermodynamic core of the ice sheet model relies on four elementary field
equations. Firstly, ice flow is assumed \emph{incompressible}. Hence, the
conservation of mass implies a conservation of volume, which can be expressed
in terms of the ice velocity vector, $\vec{v}$, by

\begin{equation}
    \label{eqn:incompressibility}
    \div{\vec{v}} = 0.
\end{equation}

Secondly, the \emph{balance of stresses} is expressed by the Stokes equation,
thus assuming creeping flow. In other words, ice is considered as a slow-moving
fluid whose deformation is entirely controlled by internal friction, here
expressed by the Cauchy stress tensor, $\bm\sigma$, and gravity, $\vec{g}$,

\begin{equation}
    \label{eqn:stressbalance}
    \tensdiv{\bm\sigma} + \rho\,\vec{g} = \vec{0},
\end{equation}

where $\rho$ corresponds to ice density. Thirdly, after defining the
strain-rate tensor
${\bm{\dot\epsilon}~=~\frac{1}{2}(\tensgrad{\vec{v}} +
                                  \tensgrad{\vec{v}}^{\mathrm{T}})}$,
and the deviatoric stress tensor, ${\bm{\tau} = \bm{\sigma} - p\,\bm{\delta}}$,
where $p=\frac{1}{3}\tr(\bm{\sigma})$ is the hydrostatic pressure and
$\bm{\delta}$ is the identity tensor, a \emph{constitutive law} for ice based
on laboratory experiments is used to relate these two quantities,

\begin{equation}
    \label{eqn:glenslaw}
    \bm{\dot\epsilon} = A\,\tau_e^{n-1}\,\bm{\tau}.
\end{equation}

The equivalent stress, $\tau_e$, is defined by
${\tau_e}^2 = -\mathrm{II}_{\bm{\tau}} = \frac{1}{2} \tr(\bm{\tau}^2)$,
where $\mathrm{II}_{\bm{\tau}}$ is the second invariant of the stress tensor.
The ice softness coefficient, $A$, depends on the deviation of temperature $T$
from the pressure-melting point $T_m$ through

@@

Letting ${\dot\epsilon_e}^2 = \frac{1}{2} \tr(\bm{\dot\epsilon}^2)$ and
$B=A^{-1/n}$, the constitutive law (\eqn{glenslaw}) can be rewritten using a
sometimes more convenient, inverted formulation,

\begin{equation}
    \bm{\tau} = B\,\dot{\epsilon}_e^{1/n-1}\,\bm{\dot\epsilon}.
\end{equation}

By analogy to Newtonian flow, an apparent viscosity, $\nu$, can then be defined
by $\bm{\tau} = \nu\,\bm{\dot\epsilon}$, thus yielding

\begin{equation}
    \label{eqn:viscosity}
    \nu = \frac{B}{2}\,\dot{\epsilon_e}^{1/n-1}.
\end{equation}

@@

\begin{equation}
    \label{eqn:enthalpy}
    \rho \left(\pdv{H}{t} + \vec{v} \cdot \grad{H} \right)
        = -\div{\vec{q}} + \frac{\nu\dot{\epsilon_e}^2}{4},
\end{equation}

where $\vec{q}$ represents the heat flow and
${\frac{\nu\dot{\epsilon_e}^2}{4} = \tr(\bm{\tau}\bm{\dot\epsilon})}$ is a
source term corresponding to strain heat release. In the case where ice
temperature, $T$, is below the pressure-melting point, the enthalpy can be
expressed as $H=c\,T$, while the heat flow is given by Fourier's law,
$\vec{q} = k\,\grad{T}$. Consequently, the enthalpy equation (\eqn{enthalpy})
can be rewritten using a more familiar, cold-ice, temperature formulation,

\begin{equation}
    \label{eqn:temperature}
    \frac{\partial T}{\partial t}
        + \vec{v} \cdot \vec{\mathrm{grad}}\,T
        = \frac{k}{\rho c} \Delta T
        + \frac{\nu\dot{\epsilon_e}^2}{4\rho c}.
\end{equation}

@@

\subsection{Shallow approximations}

@@

\begin{equation}
    \label{eqn:stressbalance}
    \tensdiv{\bm\tau} - \grad{p} + \rho\,\vec{g} = \vec{0}.
\end{equation}

@@

\begin{equation}
    \pdv{\vec{v}_{SIA}}{z} = 2A\,(\rho g)^n\,(s-z)^n\,
                             |\grad{s}|^{n-1}\,\grad{s}.
\end{equation}

@@

\begin{equation}
    \label{eqn:siassa}
    \vec{v} = \vec{v}_{SIA} + \vec{v}_{SSA}.
\end{equation}

Because $\vec{v}_{SSA}$ is small where $\vec{v}_{SIA}$ are large and
reciprocally, we can expect the model to perform well in regions where
conditions for applicability of either the SIA or the SSA are met. However,
equation~\ref{eqn:siassa} is a heuristic, and its validity in the zone of
transition between shear flow and longitudinal flow has not yet been
mathematically proven.

\subsection{Basal sliding}

@@

\begin{equation}
    \label{eqn:pseudoplastic}
    \vec{\tau}_b = -\tau_c \frac{\vec{v}_b}{{v_{th}}^q\,|\vec{v}_b|^{1-q}},
\end{equation}

where $\vec{\tau}_b$ is the basal traction force, $\vec{v}_b$ is the basal
velocity, and $v_{th}$ is an arbitrary velocity threshold. In the case of
$q=0$, equation~\ref{eqn:pseudoplastic} corresponds to a purely plastic law,

\begin{equation}
    \label{eqn:plastic}
    \vec{\tau}_b = -\tau_c \frac{\vec{v}_b}{|\vec{v}_b|}.
\end{equation}

@@

\subsection{Bedrock response}

@@

where $\nu_m$ is mantle viscosity, $\rho_l$ is lithosphere density, $D$ is the
lithosphere's flexural rigidity, and $\sigma_{zzb}$ corresponds to the ice load
\citep{Bueler.etal.2007}. Here, $\Delta^2$ designs the biharmonic operator of
linear elastic theory, while $|\grad|$ is a pseudo-differential operator
defined through the Fourier transform by \citet[Eqn.~6]{Bueler.etal.2007}. In
the left-hand part of this equation, the first term accounts for mantle
relaxation, the second for point-wise isostasy, and the third for elastic
flexure. Due to high mantle viscosity $\nu_m$, there exists a time lag between
the ice sheet growth and the isostatic bedrock response.
%$\nu=1\times10^{21}$\,Pa\,s
%$D=5\times10^{24}$\,N\,m
%$\rho_r = 3300$\,kg\,m$^{-3}$

\subsection{Surface mass balance}
\subsection{Atmospheric forcing}
\subsection{Numerical implementation}

% ----------------------------------------------------------------------
\section{Software tools and contributions}
% ----------------------------------------------------------------------
% Computational workflow
% Contributions to PISM
% PyPDD
% pismplotlib
% r.in.worldclim
% r.out.pism

@@

% ----------------------------------------------------------------------
\section{Results summary}
% ----------------------------------------------------------------------
% Constant-climate runs
% Effect of daily temperature variability
% Transient-climate runs

@@

% ----------------------------------------------------------------------
\section{Future perpectives}
% ----------------------------------------------------------------------
% Model improvements
% Fieldwork
% Geomorphological-oriented plots

>>>>

% kappa.tex
% ======================================================================

@@

% My commands
%\newcommand{\note}[1]{\textbf{[NOTE: #1]}}
\newcommand{\todo}[1]{\emph{[\textbf{Todo:} #1]}}
%\newcommand{\aref}[0]{\textbf{[ref.]}}

% Vectors and tensors
\newcommand{\vect}[1]{\va*{#1}} % bold arrow vectors
\newcommand{\tens}[1]{\vb*{#1}} % bold tensors

% Differential operators
\renewcommand{\div}[1]{\mathrm{div}\,#1}            % divergence
\renewcommand{\grad}[1]{\vect{\mathrm{grad}}\,#1}   % gradient
\newcommand{\tdiv}[1]{\vect{\mathrm{div}}\,#1}      % tensor divergence
\newcommand{\tgrad}[1]{\tens{\mathrm{grad}}\,#1}    % tensor gradient
\newcommand{\matdv}[1]{\pdv{#1}{t}+\vect{v}\cdot\grad{}\,#1}  % material dv.

% Common notations
\newcommand{\doteps}[0]{\dot{\epsilon}} % epsilon dot
\newcommand{\IDT}[0]{\tens{\delta}}     % Identity tensor
\newcommand{\CST}[0]{\tens{\sigma}}     % Cauchy stress tensor
\newcommand{\DST}[0]{\tens{\tau}}       % deviatoric stress tensor
\newcommand{\SRT}[0]{\tens{\doteps}}    % strain-rate tensor
\newcommand{\vv}[0]{\vect{v}}           % velocity vector
\newcommand{\vsia}[0]{\vv_{\mathrm{SIA}}}   % SIA velocity
\newcommand{\vssa}[0]{\vv_{\mathrm{SSA}}}   % SSA velocity

@@

% ======================================================================
\begin{document}
% ======================================================================

@@

% ======================================================================
\section{Introduction}
% ======================================================================
% Ice as viscous matter
% Glaciers and ice sheets
% The last glacial cycle

@@

% ======================================================================
\section{Field area description}
% ======================================================================
% Geographic setting
% Climatic setting
% Palaeo-glaciology

@@

% ======================================================================
\section{Numerical ice sheet model}
% ======================================================================

% ----------------------------------------------------------------------
\subsection{Overview}
% ----------------------------------------------------------------------

@@

% ----------------------------------------------------------------------
\subsection{Field equations}
% ----------------------------------------------------------------------

The thermodynamic core of the ice sheet model relies on four elementary field
equations. Firstly, ice flow is assumed \emph{incompressible}. Hence, the
conservation of mass implies a conservation of volume, which can be expressed
in terms of the ice velocity vector, $\vv$, by

\begin{equation}
    \label{eqn:incompressibility}
    \div{\vv} = 0.
\end{equation}

Secondly, the \emph{balance of stresses} is expressed by the Stokes equation,
thus assuming creeping flow. In other words, ice is considered as a slow-moving
fluid whose deformation is entirely controlled by internal friction, here
expressed by the Cauchy stress tensor, $\CST$, and gravity, $\vect{g}$,

\begin{equation}
    \label{eqn:stressbalance}
    \tdiv{\CST} + \rho\,\vect{g} = \vect{0},
\end{equation}

where $\rho$ corresponds to ice density. Thirdly, after defining the
strain-rate tensor
${\SRT = \frac{1}{2}(\tgrad{\vv} + \tgrad{\vv}^{\mathrm{T}})}$,
and the deviatoric stress tensor, ${\DST = \CST - p\,\IDT}$,
where $p=\frac{1}{3}\tr(\CST)$ is the hydrostatic pressure and
$\tens{\delta}$ is the identity tensor, a \emph{constitutive law} for ice based
on laboratory experiments is used to relate these two quantities,

\begin{equation}
    \label{eqn:glenslaw}
    \SRT = A\,\tau_e^{n-1}\,\DST.
\end{equation}

The equivalent stress, $\tau_e$, is defined by
${\tau_e}^2 = -\mathrm{II}_{\DST} = \frac{1}{2} \tr(\DST^2)$,
where $\mathrm{II}_{\DST}$ is the second invariant of the stress tensor.
The ice softness coefficient, $A$, depends on the deviation of temperature $T$
from the pressure-melting point $T_m$ through

@@

Letting ${\doteps_e}^2 = \frac{1}{2} \tr(\SRT^2)$ and
$B=A^{-1/n}$, the constitutive law (\eqn{glenslaw}) can be rewritten using a
sometimes more convenient, inverted formulation,

\begin{equation}
    \DST = B\,\doteps_e^{1/n-1}\,\SRT.
\end{equation}

By analogy to Newtonian flow, an apparent viscosity, $\nu$, can then be defined
by $\DST = \nu\,\SRT$, thus yielding

\begin{equation}
    \label{eqn:viscosity}
    \nu = \frac{B}{2}\,\doteps_e^{1/n-1}.
\end{equation}

@@

\begin{equation}
    \label{eqn:enthalpy}
    \rho \left(\matdv{H}\right)
        = -\div{\vect{q}} + \frac{\nu\doteps_e^2}{4},
\end{equation}

where $\vect{q}$ represents the heat flow and
${\frac{\nu\dot{\epsilon_e}^2}{4} = \tr(\DST\SRT)}$ is a
source term corresponding to strain heat release. In the case where ice
temperature, $T$, is below the pressure-melting point, the enthalpy can be
expressed as $H=c\,T$, while the heat flow is given by Fourier's law,
$\vect{q} = k\,\grad{T}$. Consequently, the enthalpy equation (\eqn{enthalpy})
can be rewritten using a more familiar, cold-ice, temperature formulation,

\begin{equation}
    \label{eqn:temperature}
    \matdv{T} = \frac{k}{\rho c} \Delta T
                + \frac{\nu\doteps_e^2}{4\rho c}.
\end{equation}

@@

% ----------------------------------------------------------------------
\subsection{Shallow approximations}
% ----------------------------------------------------------------------

@@

\begin{equation}
    \label{eqn:stressbalance}
    \tdiv{\DST} - \grad{p} + \rho\,\vect{g} = \vect{0}.
\end{equation}

@@

\begin{equation}
    \pdv{\vsia}{z} = 2A\,(\rho g)^n\,(s-z)^n\,|\grad{s}|^{n-1}\,\grad{s}.
\end{equation}

@@

\begin{equation}
    \label{eqn:siassa}
    \vv = \vsia + \vssa.
\end{equation}

Because $\vsia$ is small where $\vssa$ is large and
reciprocally, we can expect the model to perform well in regions where
conditions for applicability of either the SIA or the SSA are met. However,
equation~\ref{eqn:siassa} is a heuristic, and its validity in the zone of
transition between shear flow and longitudinal flow has not yet been
mathematically proven.

% ----------------------------------------------------------------------
\subsection{Basal sliding}
% ----------------------------------------------------------------------

@@

\begin{equation}
    \label{eqn:pseudoplastic}
    \vect{\tau}_b = -\tau_c \frac{\vv_b}{{v_{th}}^q\,|\vv_b|^{1-q}},
\end{equation}

where $\vect{\tau}_b$ is the basal traction force, $\vv_b$ is the basal
velocity, and $v_{th}$ is an arbitrary velocity threshold. In the case of
$q=0$, equation~\ref{eqn:pseudoplastic} corresponds to a purely plastic law,

\begin{equation}
    \label{eqn:plastic}
    \vect{\tau}_b = -\tau_c \frac{\vv_b}{|\vv_b|}.
\end{equation}

@@

% ----------------------------------------------------------------------
\subsection{Bedrock response}
% ----------------------------------------------------------------------

@@

where $\nu_m$ is mantle viscosity, $\rho_l$ is lithosphere density, $D$ is the
lithosphere's flexural rigidity, and $\sigma_{zzb}$ corresponds to the ice load
\citep{Bueler.etal.2007}. Here, $\Delta^2$ designs the biharmonic (Laplacian
square) operator of
linear elastic theory, while $|\grad|$ is a pseudo-differential operator
defined through the Fourier transform by \citet[Eqn.~6]{Bueler.etal.2007}. In
the left-hand part of this equation, the first term accounts for mantle
relaxation, the second for point-wise isostasy, and the third for elastic
flexure. Due to high mantle viscosity $\nu_m$, there exists a time lag between
the ice sheet growth and the isostatic bedrock response.
%$\nu=1\times10^{21}$\,Pa\,s
%$D=5\times10^{24}$\,N\,m
%$\rho_r = 3300$\,kg\,m$^{-3}$

% ----------------------------------------------------------------------
\subsection{Surface mass balance}
% ----------------------------------------------------------------------

% ----------------------------------------------------------------------
\subsection{Atmospheric forcing}
% ----------------------------------------------------------------------

% ----------------------------------------------------------------------
\subsection{Numerical implementation}
% ----------------------------------------------------------------------

% ======================================================================
\section{Software tools and contributions}
% ======================================================================
% Computational workflow
% Contributions to PISM
% PyPDD
% pismplotlib
% r.in.worldclim
% r.out.pism

@@

% ======================================================================
\section{Results summary}
% ======================================================================
% Constant-climate runs
% Effect of daily temperature variability
% Transient-climate runs

@@

% ======================================================================
\section{Future perpectives}
% ======================================================================
% Model improvements
% Fieldwork
% Geomorphological-oriented plots
